---
title: "Compound identification in metabolomics: an R approach"
author: "Michał Burdukiewicz"
date: "25-06-2021"
output: 
  ioslides_presentation:
    theme: "simplex"
---

<link href="https://fonts.googleapis.com/css?family=Montserrat&display=swap" rel="stylesheet">

<style>
slides > slide {
  font-family: 'Montserrat', sans-serif;
}

.center {
  display: block;
  margin-left: auto;
  margin-right: auto;

}


</style>


```{r setup, include=FALSE}
library(knitr)
library(rgl)
library(ggplot2)
library(plotly)
library(dplyr)
library(patchwork)
library(ggthemes)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(scipen=999)  # turn-off scientific notation like 1e+48
theme_set(theme_bw())  # pre-set the bw theme.
```

# Compound identification

## Summary

```{r}
p_main <- expand.grid(x = c("From compound to spectra\n", "From spectra to compound\n"),
            y = c("Quantum chemistry", "Combinatorial optimization methods",
                  "Heuristic-based methods", "Machine learning models")) %>% 
  mutate(lab = c("Small compounds,\nmechanistic insights", "", 
                 "Small compounds, lipids", "Small compounds,\nlipids",
                 "Lipids", "Lipids",
                 "Small compounds", "Small compounds,\nbiotransformations")) %>% 
  ggplot(aes(x = x, y = y, label = lab)) +
  geom_tile(fill = NA, color = "black") +
  geom_text() +
  theme_void() +
  theme(axis.text = element_text(color = "black"))

p_main
```

## Software

```{r}
p_main <- expand.grid(x = c("From compound to spectra\n", "From spectra to compound\n"),
            y = c("Quantum chemistry", "Combinatorial optimization methods",
                  "Heuristic-based methods", "Machine learning models")) %>% 
  mutate(lab = c("QCEIMS", "", 
                 "MIDAS-G, MetFrag, MAGMa+", "SIRIUS",
                 "Mass Frontier, LipidBlast,\nLipidMatch", "Lipid Data Analyzer, LipidXplorer",
                 "CFM-ID., NEIMS", "FingerID, DeepEI")) %>% 
  ggplot(aes(x = x, y = y, label = lab)) +
  geom_tile(fill = NA, color = "black") +
  geom_text() +
  theme_void() +
  theme(axis.text = element_text(color = "black"))

p_main
```

# CI workflow

## Two-step procedure

1. Identify the proper database using CEU Mass Mediator.
2. Do a MetFrag search using an identified database.

## CEU Mass Mediator

Available for the command line processing as an R package.

Gil-de-la-Fuente, A., Godzien, J., Saugar, S., Garcia-Carmona, R., Badran, H., Wishart, D.S., Barbas, C., and Otero, A. (2018). CEU Mass Mediator 3.0: A Metabolite Annotation Tool. Journal of Proteome Research.

## CEU Mass Mediator

```{r,echo=TRUE}
library(httr)
r <- POST("http://ceumass.eps.uspceu.es/api/batch", 
          body = "Is anyone here?", encode = "json")

res <- try(stop_for_status(r), silent = TRUE)

print(res)
```


## MetFrag

Available for the command line processing as an R package cmmr.

Ruttkies, C., Schymanski, E.L., Wolf, S., Hollender, J., and Neumann, S. (2016). MetFrag relaunched: incorporating strategies beyond in silico fragmentation. Journal of Cheminformatics 8, 3.

## MetFrag 

```{r,echo=FALSE}
settingsObject<-list()
settingsObject[["DatabaseSearchRelativeMassDeviation"]]<-5.0
settingsObject[["FragmentPeakMatchAbsoluteMassDeviation"]]<-0.001
settingsObject[["FragmentPeakMatchRelativeMassDeviation"]]<-5.0
settingsObject[["MetFragDatabaseType"]]<-"PubChem"
settingsObject[["NeutralPrecursorMass"]]<-253.966126
settingsObject[["PeakList"]]<-matrix(c(
  90.97445, 681,
  106.94476, 274,
  110.02750, 110,
  115.98965, 95,
  117.98540, 384,
  124.93547, 613,
  124.99015, 146,
  125.99793, 207,
  133.95592, 777,
  143.98846, 478,
  144.99625, 352,
  146.00410, 999,
  151.94641, 962,
  160.96668, 387,
  163.00682, 782,
  172.99055, 17,
  178.95724, 678,
  178.97725, 391,
  180.97293, 999,
  196.96778, 720,
  208.96780, 999,
  236.96245, 999,
  254.97312, 999), ncol=2, byrow=TRUE)
settingsObject[["NeutralPrecursorMolecularFormula"]]<-"C7H5Cl2FN2O3"
settingsObject[["PrecursorCompoundIDs"]]<-c("50465", "57010914", "56974741", "88419651", "23354334")
#
# pre and post-processing filter
#
# define filters to filter unconnected compounds (e.g. salts)
settingsObject[["MetFragPreProcessingCandidateFilter"]]<-c("UnconnectedCompoundFilter","IsotopeFilter")
settingsObject[["MetFragPostProcessingCandidateFilter"]]<-c("InChIKeyFilter")
settingsObject[1L:5]
```

## MetFrag

```{r,echo=FALSE}
settingsObject[6]
```

## MetFrag

```{r,echo=FALSE}
settingsObject[7L:10]
```

## MetFrag - results

```{r,echo=FALSE}
knitr::kable(readRDS("./data/metfrag-result.RDS")[1L:3, 1L:3])
```

## MetFrag - results

<div style="float: left; width: 40%;">
```{r,echo=FALSE,results='asis'}
cat(paste0("* ", colnames(readRDS("./data/metfrag-result.RDS"))[1L:9]), sep = "\n\n")
```
</div>

<div style="float: right; width: 60%;">
```{r,echo=FALSE,results='asis'}
cat(paste0("* ", colnames(readRDS("./data/metfrag-result.RDS"))[10L:18]), sep = "\n\n")
```
</div>

# Summary

## General remarks

1. We are not limited to R.
2. Intented solution won't be a command line tool, but rather a clickable interface.

## Questions

1. Why run CEU Mass Mediator at all? Instead, we can run MetFrag on all available databases.
2. What is the easiest way to import data? How to limit the number of clicks?
3. Can we make a solution independent from the web services?



